---
title: "BycatchEstimatorStan User Guide"
author: "Beth Babcock"
date: "`r Sys.Date()`"
output: html_document
---

This library runs a Bayesian version of the model-based bycatch estimation procudure in the library BycatchEstimator. See the BycatchEstimator User's Guide for how to set up the data at https://ebabcock.github.io/BycatchEstimator/. The source code is at https://github.com/ebabcock/BycatchEstimator

To install, use the following code:

```{r}
# install.packages("devtools")
#devtools::install_github("ebabcock/BycatchEstimator")
#devtools::install_github("ebabcock/BycatchEstimatorStan")
evalOption<-FALSE
```

Open all the libraries

```{r, eval = evalOption}
library(BycatchEstimator)
library(MuMIn)
library(BycatchEstimatorStan)
# library(gridExtra)
# library(rstan)
# library(cmdstanr)
# library(loo)
# library(shinystan)
# library(ggmcmc)
# library(readxl)
# library(bayesplot)
# library(ggsci)
# library(flextable)
theme_set(theme_bw())

#source("~/github/bycatchEstimatorStan/R/printStanCode.r")
#source("~/github/bycatchEstimatorStan/R/bycatchStan.r")
```

Run example code from BycatchEstimator

```{r, eval = evalOption}
obsdat<-droplevels(LLSIM_BUM_Example_observer[LLSIM_BUM_Example_observer$Year>2014 &LLSIM_BUM_Example_observer$fleet==2,])
logdat<-droplevels(LLSIM_BUM_Example_logbook[LLSIM_BUM_Example_logbook$Year>2014 & LLSIM_BUM_Example_logbook$fleet==2,])
setupObjBUM<-bycatchSetup(
  obsdat = obsdat,
  logdat = logdat,
  yearVar = "Year",
  obsEffort = "hooks",
  logEffort = "hooks",
  factorVariables = c("Year","area"),  
  numericVariables = NA, 
  EstimateBycatch = TRUE,
  logNum = NA,
  sampleUnit = "trips",
  baseDir = getwd(),
  runName = "LLSIMBUMtripExample",
  runDescription = "LLSIm BUM by trip, with 5% observer coverage",
  common = "Blue marlin",
  sp = "Makaira nigricans",
  obsCatch ="BUM",
  catchUnit = "number",
  catchType = "catch",
  reportType = "html"  
)
```

Look at the output html file to see whether there are any problems with your data setup. 

Now run BycatchStan

```{r, eval = evalOption}
modelsToRun<-c("y~Year","y~1")
useCode="cmdstanr"
BUMRun<-bycatchStanSim(setupObjBUM,
                        modelsToRun=modelsToRun,
                        spNum=1,  #which species to run in input is multispecies
                        modeledEffort=FALSE,
                        priors =  list(interceptSD=10,
                                      coefficientSD=1,
                                      phiType=c("exponential","normal")[1],
                                      phiPar=1),
                        StanOutDir=NULL,  #Null to use base directory from SetupObj
                        useCode=useCode)

```

The bycatchStanSim function prints its output to and rds file, so you can read it back in later to make plots and look at diagnostics. 

```{r, eval = evalOption}
#Specify directory with bycatch Estimator results
outDir<-setupObjBUM$bycatchInputs$outDir
# Date of bycatch estimator run
estimatorDate<-Sys.Date()
# Date of stan run
stanDate<-Sys.Date()
# Read in bycatchEstimator setup
setupObj<-readRDS(paste0(outDir,"/",estimatorDate,"_BycatchSetupSpecification.rds"))
spNum<-1
#Stan output directory
StanOutDir<-paste0(outDir,"/","Bluemarlincatch")
# Read in stan run summary files
stanSum<-readRDS(paste0(StanOutDir,
                         "/",stanDate,"StanOutputs.rds"))
```

The following code takes either the output that was just read in or the outputs from this section and looks at more diagnostics.

```{r, eval = evalOption}
#Choose outputs to use
stanSum<-BUMRun
setupObj<-setupObjBUM
```

Plot the annual bycatch estimates

```{r, eval = evalOption}
plotStan(stanSum$yearSum)
```

Here is the model selection table

```{r, eval = evalOption}
stanSum$waictab
```

Pick AIC best model

```{r, eval = evalOption}
modelNum<-which(stanSum$waictab$Dwaic==0) #To get WAIC best
```

Read in the corresponding Stan model object (these are deleted from the environment to save space)

```{r, eval = evalOption}
stanObj<-readRDS(BUMRun$stanRunFiles[[modelNum]])
```

Plot residuals
```{r, eval = evalOption}
getResiduals(stanSum, 
             stanObj,
             modelNum,
             setupObj,
             useCode=useCode)
```

Plot prior and posterior densities of parameters

```{r, eval = evalOption}
plotPriorPosterior(stanSum, 
                   stanObj,
                   useCode=useCode) 
```

Plot prior and posterior draws of bycatch 

```{r, eval = evalOption}
plotPriorPosteriorSims(stanSum,
                       modelNum,
                       stanObj,
                       setupObj,
                       modeledEffort = FALSE,
                       effortSD = NULL,
                       useCode=useCode)  
```


Convergence diagnostics

```{r, eval = evalOption}
getConvergence(stanObj,
               useCode=useCode)
```

Parameter summary table

```{r, eval = evalOption}
getSummary(stanSum,
           stanObj,
           modelNum,
           setupObj,
           spNum=1,
           useCode=useCode) 
```

## Code to run mortality model 


With simulated survival data (arbitrary values)

```{r, eval = evalOption}
mortData<-read.csv("data/SimulatedMortality.csv")
mortData<-standardizeToObsdat(obsdat,mortData,numericVariables="Year")
#with no prediction
modelsToRun=c("y~1","y~Species","y~Year","y~Year+Species","y~Year:Species")
MortResults<-mortalityStan(mortData=mortData,
                          predData = NULL,
                          modelsToRun=modelsToRun,
                          aliveColumn="alive",
                          outDir=getwd(),
                          runName="Simulated",
                          predictP=FALSE,  
                          useCode=useCode  
)
```

The outputs are in a similar format.

```{r, eval = evalOption}
MortResults$waictab
MortResults$waictab
MortResults$diagTable
mortModelNum<-which(MortResults$waictab$Dwaic==0)
```

With predictions for each year for BUM, best model only

```{r, eval = evalOption}
predData<-obsdat %>%
  mutate(Species="BUM",
         Species=factor("BUM",levels=c("BUM","SWO"))) %>%
  distinct(Species,Year)
dim(predData)
predData
MortResultsP<-mortalityStan(mortData=mortData,
                           predData = predData,
                           modelsToRun=modelsToRun[mortModelNum],
                           aliveColumn="alive",
                           outDir=getwd(),
                           runName="Simulated",
                           predictP=TRUE,  
                           useCode=useCode  
)
```

## Combine mortality and bycatch estimate to get bycatch mortality
Not working yeat

```{r, eval = evalOption}
mortalityEsts<-getMortPred(stanSum=stanSum,
                           modelNum=which(stanSum$waictab$Dwaic==0),
                           stanObj=stanObj,
                           Species="BUM",
                           setupObj=setupObj,
                           modeledEffort = FALSE,
                           effortSD = NULL,
                           useCode=useCode,
                           mortResults=MortResultsP,
                           mortModelNum=which(MortResultsP$waictab$Dwaic==0))
plotMortalityFunc(mortalityEsts,Species="BUM")
head(mortalityEsts)
```
